\chapter{Artificial intelligence in drug discovery}

\section{Artificial intelligence in drug discovery - part 1}

	\subsection{Context}
	The application of algorithms in drug discovery needs at its core to consider which data are being analysed and which methods are used for this purpose.
	The major developments that have been proposed for this are:

	\begin{multicols}{2}
		\begin{enumerate}
			\item Flowering of computational thermodynamics.
			\item Learning to turn potent ligands into drug candidates.
			\item New classes of drug targets which will challenge the competencies.
			\item Encountering novel molecular mechanism for drug efficacy.
			\item Leaning to model an entire signal transduction pathway and use that to better select drug targets.
			\item The technology will disperse.
			\item Virtual screening will become routine.
		\end{enumerate}
	\end{multicols}

	Most of this has become reality.
	In vivo pharmacokinetics and metabolism are still tricky to deal with and modelling entire signal transduction pathways is possible on a local basis, but the full impact is difficult to ascertain.
	Giving these tools to medicinal chemists is a challenge due to the difference in thinking and approaches.
	The difficult part of current computational drug-discovery is the understanding of more complex biology.
	This is because on a biological system is more difficult to define a finite set of parameter and so there is more uncertainty.
	We are unable to determine which variables matter, define them experimentally and label the biology for AI to succeed on the level that is compatible with the current investment and hope in the area.

	\subsection{Quality versus speed}
	It can be seen that a reduction of the failure rate has the most significant impact on project value overall, more than reduction of cost of a phase or a decrease in the amount of time of a pahse.
	This has the most impact in the clinical phase.
	AI can make decision faster and more cheaply and lead to better decision.
	In the context of drug discovery the last aspect will have the biggest impact.
	So for drug-discovery AI needs to support:

	\begin{multicols}{2}
		\begin{enumerate}
			\item Better compounds going into clinical trial.
			\item Better validated targets.
			\item Better patient selection.
			\item Better conductance of trials.
		\end{enumerate}
	\end{multicols}

	For AI to show its value in drug discovery the focus needs to shift from the early stage with Pearson correlation and RMSE.
	This is so the sole focus of models is not only on incrementally improving numerical values obtained fr models or proxy measures of success.

	\subsection{Current analytical way of drug discovery using isolated mechanisms and targets}
	In biology deficiencies in one or more of the components are the cause of disease and their modulation will allow to cure it.
	SO this is the process of drug discovery via isolated model systems.
	This is limited due to the connectivity of biology with dynamics often unknown, insufficiently formalized and quantified.
	Phenotypic screening has attempted to merge disease-relevant biology with large numbers of compounds that can be screened.
	This comprises a lot of hypothesis-driven and hypothesis-free readouts.
	If the former is true the argument for a phenotypic screen is decreased and the understanding of the high-dimensional readout can be non-trivial.
	Target identification is still a problem, so phenotypic screening incorporate some biological complexity, but it is not able to recapitulate intracellular organ toxicity to a reasonable extent.
	The basic principle of modern drug discovery is to identify a protein that does not function as needed in a diseased system and find modulators of that mechanism.
	This doesn't deal with the undesired effects and interaction that this molecule can have in vivo.
	This approach is plausible only in the case of monocausal diseases.
	Also achieving activity in a model system neglects the question of whether the compound reaches its intended target site, it is able to revert the diseaed phenotype and it achieves the goal with tolerable side effects.
	Many current approaches of AI aim to transfer methods from image or speech recognition with the aim of improving the prediction of a particular molecular property end point quantitatively.

	\subsection{Chemical and biological properties of relevance for drug discovery and the extent to which they are captured in current data}

	\begin{table}[H]
		\centering
		\begin{tabular}{|c|c|c|}
			\hline
			Assays & Comments & Useful as predictor for human in vivo situation?\\
			\hline
			\makecell{On-target activity \\as a proxy for\\efficacy} & \makecell{Target validation \\difficult, weak links between\\individual targets\\ and phenotypic effects} & \makecell{Target engagement in vivo\\ is frequently sufficient for causing\\phenotypic effects, however\\ activity on a target in isolated\\protein or cellular assays\\ often differs from the situation in vivo}\\
			\hline
			\makecell{Physiochmical\\properties} & \makecell{Relevant for almost\\any aspect of drug\\development owing to\\the application of drugs in\\largely aqueous biological\\systems} & \makecell{Crucial especially for orally\\administered drugs, broad\\correlation with many drug\\properties, especially in ADME space\\needs to be considered\\in combination with\\other compound properties}\\
			\hline
			\makecell{PK and points} & \makecell{Generally simplified cellular\\systems to anticipate\\uptake, metabolism and\\transport of compounds\\across body compartments.\\Simplified version} & \makecell{Organs and tissues\\more complex and\\heterogeneous. Adaptive\\responses and impact\\of the microbiome\\neglected}\\
			\hline
			\makecell{Cellular\\toxicity} & \makecell{Human HepG2 surrogate\\effects of toxicity\\in liver,\\cytotoxicity early stage\\marker for adverse effects of\\compounds} & \makecell{Heterogeneity\\not considered}\\
			\hline
			\makecell{Heterogeneous cell\\cultures} & \makecell{Resemble organ systems\\better} & \makecell{More representative\\more difficulte to handle}\\
			\hline
			\makecell{Toxicity\\Safety} & \makecell{Many proxy end points\\ on individual readouts} & \makecell{Therapeutic index\\difference in exposure between\\efficacy and safety.\\Estimate organ based\\toxicity in\\quantitative way}\\
			\hline
			\makecell{Animal models} & \makecell{Representitativeness\\and predictivity\\not a given} & \makecell{Depends on the case}\\
			\hline
		\end{tabular}
	\end{table}

	\subsection{Current landscape}

		\subsubsection{De novo design}
		In de novo design there is forward prediction and retrosynthesis prediction.
		They aim to identift how chemical matter can be synthesized.

		\subsubsection{Docking}
		Docking deals with discerning if the ligand, once in place, binds to a target.

		\subsubsection{Connection from mode of action to phenotypic effect}
		Connection with genetic support and functional genomics.

		\subsubsection{Integration}
		Integrating ligand-protein activity and target identification and PC properties of the compound in an integrated matter is still missing and would involge modelling the interactions of a small molecule with all of its interaction partners, addressing the question of target expression in the disease tissue and its involvement of disease modulation, including the PK behaviour of a molecule with respect to the in vivo system and considering safety in parallel with efficacy.

	\subsection{Validation of AI models in drug discovery}
	If AI approaches for drug discovery only end up generating a ligand for a protein, then there is no evidence that this will help drug discovery as a whole.
	There is a need to move to more complex biologica systems earlier.
	This means including more predictive end points in models.
	There are no control experiments being conducted when AI delivers new compounds.
	Given that a drug coming to marker is a long series of choice it is impossible to disentangle whether the end product is a result of the method applied or the result of subjective choices on which compound to test.
	The large chemical space tends to lead to trivial validation examples, also method validation in the chemical domain is difficult.
	One is never able to truly prospectively test models, except with a very large prospective experiments and controls.
	The performance reported and the data are related.
	Comparative benchmarking datasets are retrospective but a true estimate of model performance is unobtainable.
	Also there is a need to develop algorithms considering the medical domain to avoid biases and to be able to handle biological drifts.

\section{Artificial intelligence in drug discovery - part 2}

	\subsection{Introduction}
	Datasets used in biological, chemical and drug discovery domains have little data compared to others in the order of hundreds for in vivo annotations.
	In these domains the situation of how to represent information is difficult.
	In the chemical domain it is not clear which description captures the most relevant informations.
	Some molecular properties are defined by local feature, whereas other by a distribution of properties across the molecular surface or on external context.
	There are a lot of ways to represent molecules, but there is limited understanding of which representations are predictive of the property endpoint of interest.
	It is unknown what matters about a structure for a given effect.
	For example SMILE tables that gives information about the connectivity of a molecule does not give information about its spatial features.
	Endpoints can be modelled in a meaningful way using connectivity table-derived descriptors where sufficient data for a relevant endpoint are available to truly learn representations.
	This is not the cased compared with the size of chemical space.
	Also the label depends and it is not defined entirely by the compound structure.
	Similar considerations apply on a larger scale on biological data.
	It is not trivial to understand the relationships of different levels of biology with particular endpoints of interest.
	Also the dynamics nature of the environment poses problems.
	SO for the chemical and biological data available the problems are due to the amounts of data available, the ability to represent it in a suitable form to the computer, a ML algorithm that is intrinsically consistent with the data avaiable and the possibility to assign meaningful labels.
	In the biological domain system do not follow explicitly defined rules or they are at least not known and can be defined on a large number of different levels.
	Observation are highly conditionals and the parameters they depend on are not known.

	\subsection{Why representing drug discovery information for AI is difficult: linking bioactivity to adverse events}
	Although one might assume that activity against a protein target is sufficient to understand its effect in a biological system this is not the case in practice.
	Considering the ration of bioactivity against a target relative to unbound plasma concentration, with respect to the likelihood of a drug showing adverse events for annotations.
	The hypothesis is that if the blood plasma concentration of a drug is above the threshold needed to act on a particular target, then one would assume to see a particular type of side effects.
	Single descriptors of a compound will not be able to anticipate the full biological complexity of drug effects.

	\subsection{Data and problem settings in drug discovery}
	For the purpose of data analysis methods in the drug discovery domain, two types of model can be distinguished: those based on large-scale and proxy measures, which aim at compound selection from a large set of virtual molecules and which tend to be qualitative in nature, and models that are based on the available data on a smaller scale with a more quantitative nature and aim for the prediction of safety or efficacy endpoints.
	Drug discovery often deals with proxy measures where large numbers of data points can be generated by simple assays.
	In cases where this measures have low productivity the amount of data is not sufficient to generate a useful model.
	Quantitative readouts from complex biology might be more advantageous.
	There is no general mapping of proxy endpoints to a particular efficacy or toxicity.
	Currently there is not the data of the right types available to enable the generation of models to utilize AI for drug discovery.
	The understanding of biology is insufficient to define what the measured is needed.

	\subsection{Data types and their information content in drug discovery}
	Ontologies or hierarchical labelling concepts are used in the biological and drug discovery fields, with an helpful way to organize chemical and biological information.
	This annotation are far from ideal: for example mode of action is not sufficient because there are many different ways of interacting with a drug target and only a few of them are functionally equivalent to each other.
	Also different type of interaction are possible.
	This inability to label compounds in a meaningful way leads to problems when mining data.
